---
title: "Introduction 2 R"
author: "Joel Parker"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
link-citations: yes
description: "This is an intro to R workshop."

---

# The Prerequisites
This is an intro to R workshop for coders with little exirence. Of course all levels or R users are welcome to attend and engage in the content. In this work shop we will learn basics of the tidyverse packages.   \newline

The material covered in this workshop closely follows the 2020 ResBaz intro to R course found here https://picoral.github.io/resbaz_intro_to_r/index.html


## Why use R and R Studio

### More coding, and less point and click
The results in R rely on a series of well defined commands rather than just pointing and clicking. This can allow you to have a better understanding and control of the results. The steps taken in the analysis can be saved making it fast to run the analysis again. This makes statisitical analysis easily reproducible, which is important when conducting research. 

## Wrote by the people for the people
There are 10,000 + packages that can be used in R. Packages are easy to download and implement. If you can't find a package for your specific need, then you can create your own for yourself and others to use. 

## R has a large community

Thousands of people use R daily. Many of them are willing to help you through mailing lists and websites such as Stack Overflow.


## R produces high-quality graphics
packages like ggplot allow users to create high quality graphics like the one below. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(Lahman)
data(Teams)

## filter data to just Cubs and Indians since both were playing, in 1901
## Teams didn't play the same number of games each year, so rescale
## Use team names instead of codes, and clean up unused teams
library(dplyr)
library(forcats)
library(magrittr)
Teams %>%
 filter(franchID %in% c("LAD", "STL", "MIN") & yearID > 1900) %>%
 mutate(Team = as.character(franchID)) %>%
 mutate(RunsPerGame = R / G) %>%
 mutate(HitsPerGame = HA / G) -> CardsDodg.team






  ggplot(data = CardsDodg.team, aes(yearID,RunsPerGame)) +
  theme_bw()+
    geom_point(aes(color =Team)) +
    geom_smooth(aes(color = Team), method = "loess") +
    xlab("Year") +
    ylab("Runs per game") +
    ggtitle("Runs per Game") +
    theme(plot.title = element_text(hjust = 0.5)) +
    scale_color_manual(values =  c("blue","darkgreen","red"))










```


## Open Source
Anyone can inspect the source code to see how R works. Because of this transparency, there is less chance for mistakes, and if you (or someone else) find some, you can report and fix bugs.


# Installing R and R Studio
If you are running your R code in your computer, you need to install both R and RStudio. Alternatively, you can create a free account at http://rstudio.cloud and run your R code in the cloud. Either way, we will be using the same IDE (i.e., RStudio).

<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
<div class = "blue">

What’s an IDE? IDE stands for integrated development environment, and its goal is to facilitate coding by integrating a text editor, a console and other tools into one window.

</div>

##  Installing R and R studio

1. Download and install R from https://cran.r-project.org
2. Download and install RStudio from https://rstudio.com/products/rstudio/download/#download

## I already have R and RStudio installed.
1. Open RStudio
2. Check your R version by entering sessionInfo() on your console.
3. The latest release for R was April 24, 2020 (R version 4.0.0, 2020-04-24, Arbor Day). If your R version is older than the most recent version, please follow step 1 in the previous section to update R.
4. Check your RStudio version, if your version is older than Version 1.2.5042, please follow step 2 in the previous section to update RStudio.


# Intro to R Part II
## Install tidyverse
For this workshop we are going to be using packages from the tidyverse. More information on tidyverse can be found here https://www.tidyverse.org/
 \newline

To start download the tidyverse packages using install.packages()
```{r eval=FALSE}
#install tidyverse
install.packages("tidyverse")

```


Now load the libraries
```{r}
# Load tidyverse
library(tidyverse)

```

Notice all of the different packages that are not available for use to use. 

* ggplot2
* tibble
* tidyr
* readr
* purr
* dplyr
* stringr
* forcats

We will mostly be using the dplyr and ggplot packages.  

## Loading Data


For this workshop we are going to be using data from the Lahman baseball database. To install the Lahman package use install.packages("Lahmen")
```{r eval=FALSE}
# install lahmen data set. 
install.packages("Lahman")

```

and load into your enviroment. 

```{r}
library(Lahman)
```



# dplyr

dplyr has many usefull functions and it is not nessisary to remember all of the functions. Cheat sheets have be created to help us we get stuck. The dplyr cheat sheet can be found here: https://github.com/rstudio/cheatsheets/blob/master/data-transformation.pdf

## Batting data. 
The Lahmen database consists of many datasets. For this we will be using the teams data set. You can load this into your enviroment with the data function. 

```{r}
data(Teams)

```


Check your R enviroment. You should see the Teams data loaded. \newline

Use the head function to see the first 10 rows of data.

```{r}

# View the first 10 rows. 
head(Teams,n=10)

```


## filter
Typically we what are called pipes %>% to edit and manipulate data. We will start by filtering data. Lets say we were only interested in the data that was collected after 1950. We can use the filter function from dplyr to get what we want. 

```{r}

# Check the class of the year variable
class(Teams$yearID)

# year is stored as an integer so we will need to use integer values to filter. 



# data after 1950

Teams %>%
  filter(yearID > 1950)

```



You can use the &  or | opperators if multiple conditions need to be met. Lets use the filter functions to meet the following conditions

* yearID after 1950
* franchID is equal to BOS (Boston)

```{r}

## Filter with multiple conditions. 
Teams %>%
  filter(yearID > 1950 & franchID =="BOS" )
```


* yearID after 1950
* franchID is equal to BOS (Boston) or LAD (Los Angelas Dodgers). 

```{r}

## Filter with multiple conditions. 
Teams %>%
  filter(yearID > 1950 & (franchID =="BOS" | franchID=="LAD"))
```



### Challenge

Use the teams data and the filter function to meet the following criteria. 

* yearID greater than 1900
* franchID equal to "LAD", "STL" or  "MIN"
* save this dataframe as plot.data. 

Hint:
```{r eval=FALSE}
plot.data <- Teams %>%
  filter()

```



## group_by
Somtimes your data may be in different groups. Lets say we are interested in how many games per year each franchise wins. We can use group_by and Summary. 

```{r}

## Summarise wins per year for each team.
Teams %>%
  group_by(franchID) %>%
  summarise(wins_per_year = mean(W))

```


## Mutate 

Many times it is nessisary to create new variables or manipulate old ones. You can use the mutate function to do this. \newline

* Lets create create a new variable call win_percentage
* Also use the select function to only look at the yearID,franchID,W,G and win_percentage 

```{r}
Teams %>%
  mutate(win_percentage = W/G) %>%
  select(yearID,franchID,W,G,win_percentage)


```


### Challenge

USing the plot.data data created in the previous challenge. Use the mutate function to create the following variables

* Team: MIN for franchID== Min, LAD for franchID=="LAD" and STL for franchID=="STL"
* RunsPerGame:  R/G
* HitsPerGame: HA/G





# Q and A
Please think about why you came to this workshop and what you were hoping to get out of it. What are you planning to use R for in the future? Lets have an open discussion about new topics we may have not disscussed yet. 






